\filestart

\section{Modal logic}\label{ModalChapter}
\subsection{Preview}
In this chapter, we will show how to interpret the $\Box$ operator of traditional modal logic in the context of introspective theories (or their generalizations). In particular, after defining the $\Box$ operator in this context, we observe in this chapter how it satisfies the rules of the modal logic K4.

\subsection{The box operator}\label{BoxOperatorDefn}
The following notation will be very convenient for us going forward. It is also suggestive of connections with modal logic we will eventually explore:

Let $\langle T, C \rangle$ be a locally introspective theory.

Recall from \magicref{IndexedPresheafDefn} the $T$-indexed category $\Psh{C}$ (the appropriate notion of the category of presheaves on $C$ when $C$ is an indexed category rather than a category simpliciter).

Thus, we have three $T$-indexed lexcategories of note: $T$ itself (considered as a $T$-indexed category through the self-indexing $T/-$), $C$, and $\Psh{C}$.

Between these, we also have a cycle of $T$-indexed lexfunctors, like so:

% https://q.uiver.app/#q=WzAsMyxbMiwwLCJULy0iXSxbMCwyLCJDIl0sWzQsMiwiXFxQc2h7Q30iXSxbMCwxLCJcXGludHJvRiIsMl0sWzEsMiwiYyBcXDsgXFxtYXBzdG8gXFxIb21fQygtLCBjKSIsMl0sWzIsMCwiUCBcXG1hcHN0byBQKDEpIiwyXV0=
\[\begin{tikzcd}
	&& {T/-} \\
	\\
	C &&&& {\Psh{C}}
	\arrow["\introF"', from=1-3, to=3-1]
	\arrow["{c \; \mapsto \Hom_C(-, c)}"', from=3-1, to=3-5]
	\arrow["{P \mapsto P(1)}"', from=3-5, to=1-3]
\end{tikzcd}\]

Here, the bottom arrow is the Yoneda embedding, sending each object of $C$ to the corresponding representable presheaf. The right arrow takes a presheaf on $C$ to its evaluation at the terminal object of $C$; that is, to its global elements. The left arrow is the $\introF$ which is part of the structure of an introspective theory.

\begin{observation}\label{SliceBoxIsAspectBox}
Via \magicref{AspectIsSliceGlobal}, we have the $t$-aspect of the above triangle of indexed lexfunctors is the same as the global aspect of the same triangle of indexed lexfunctors relative to the slice locally introspective theory $\slicePreIntrosp{T}{t}$ given by \magicref{IntrospSlice}. \sTODOinline{Word this better.}
\end{observation}

\begin{definition}\label{BoxDefn}
In general, we will write $\Box$ for a roundtrip around this diagram, starting from any of its three nodes.

Thus, we will write $\Box$ for the $T$-indexed lexfunctor from $T$ to itself given by $t \mapsto \Hom_C(1, \introF(t))$.

We will ALSO write $\Box$ for the $T$-indexed lexfunctor from $C$ to itself given by $c \mapsto \introF(\Hom_C(1, c))$.

And we will ALSO write $\Box$ for the $T$-indexed lexfunctor from $\Psh{C}$ to $\Psh{C}$, which sends the presheaf $P$ to the presheaf represented by $\introF(P(1))$.

When we want to clarify precisely the domain we are operating on, we will write names such as $\Box_{T/-}$, $\Box_C$, or $\Box_{\Psh{C}}$, as appropriate.

As the Yoneda embedding is naturally thought of as the inclusion of a full sub-lexcategory, identifying $C$ with the corresponding representable presheaves within $\Psh{C}$, we may also think of $\Box_{\Psh{C}}$ as a $T$-indexed lexfunctor from $\Psh{C}$ to $C$. That is, as the composition of merely the top two arrows above.
\end{definition}

The above was all discussed for $T$, $C$, and $\Psh{C}$ considered as $T$-indexed lexcategories, but this all (and the rest of this chapter as well) descends to corresponding structure on their global aspects as well. In particular, we may write $\Box_{T}$ to denote the global aspect of $\Box_{T/-}$. Keep in mind, the global aspect of $\introF$ is $\introS$, so wherever in the above we discussed $\introF$, this may be rewritten as $\introS$ when considering just the global aspect.

\bigskip
\begin{remark}\label{BoxNotationSmallnessConcerns}
The reason we restricted attention here to $T$-\repsmall/ $(T, C)$-indexed sets (these being the objects of $\Psh{C}$ as we've defined it), rather than arbitrary $(T, C)$-indexed sets, is so that the map $P \mapsto P(1)$ can indeed be taken as always landing back within $T$, and not within $\Psh{T}$ more generally.

Similarly, the reason we restricted attention to locally introspective theories (i.e., the case where $C$ is locally $T$-\repsmall/), and not to pre-introspective theories more generally, is so that the Yoneda embedding $c \mapsto \Hom_C(-, c)$ does indeed yield $T$-\repsmall/ $(T, C)$-indexed sets.
\end{remark}

\sTODOinline{
\subsection{Box preservation}
\sTODOinline{This section may not be particularly necessary. We may remove it.}

Note that, having set up these various notions of $\Box$, we find that $\Box$ and each of the maps in the diagram \quote{commutes} in the appropriate sense; that is, they can be seen as preserving each other.

For example, $\introF$ preserves $\Box$, in that both $\introF(\Box(-))$ and $\Box(\introF(-))$ yield the same $T$-indexed functor from the self-indexing of $T$ to $C$. This is readily seen by unwinding their definitions: These are both $\introF(\Hom_C(1, \introF(-)))$.

We also have that taking $\Hom_C(1, -)$ preserves $\Box$ in the same way. $\Hom_C(1, \Box(-)) = \Box(\Hom_C(1, -)) = \Hom_C(1, \introF(\Hom_C(1, -)))$. More generally, consider the map $G$ which assigns to every $P \in \Psh{C}$ its object of global elements $P(1)$. Then $G$ preserves $\Box$ in the same way: $G(\Box P) = \Box(G(P)) = \Hom_C(1, \introF(P(1)))$.

Given object $c$ of $C$, and $P \in \Psh{C}$, we will write $c \implies P$ to indicate the exponential presheaf $P^c$. That is, the presheaf $P(- \times c)$. We may also write $c \implies d$ where $d$ is an object of $C$ as well, to mean $c \implies P$ for the presheaf $P = \Hom_C(-, d)$ represented by $d$. Thus, $c \implies d$ is the presheaf $\Hom_C(- \times c, d)$.

Note that for any $P \in \Psh{C}$ and object $c$ of $C$, we have that $P(c)$ can be identified with the presheaf $c \implies P$ evaluated at $1$, and thus $\introF(P(c))$ can be identified with $\Box(c \implies P)$. In particular, in the case when $P$ is represented by object $d$, we find that $\Box(c \implies d)$ is naturally identifiable with $\introF(\Hom_C(c, d))$.
}

\subsection{Modal logic and axiom 4}\label{ModalAxiom4Section}
The choice of this $\Box$ notation for these purposes is meant to convey an analogy with the $\Box$ operator of modal logic, and in particular, with the provability operator of provability logic. We will explore this more in later remarks.

The key point here is that the rules of the $\Box$ operator in a Kripke normal modal logic are essentially the rules of a lex endofunctor on a lexcategory, and any of our $\Box$ operators is certainly lex as a composite of lex functors.

Furthermore, each of our $\Box$ operators comes with a natural transformation from $\Box$ to $\Box \Box$ corresponding to the so-called 4 axiom $\Box A \vdash \Box \Box A$ in modal logic.

For the $\Box$ operator on $T$ this is clear, as the natural transformation $\introN$ from identity to $\Box$ encodes the even stronger property $t \vdash \Box t$. The 4 axiom is the special case where $t$ here is of the form $\Box A$.

% https://q.uiver.app/?q=WzAsMyxbMiwwLCJUIl0sWzAsMiwiXFxHbG9ie0N9Il0sWzQsMiwiXFxHbG9ie1xcUHNoe0N9fSJdLFswLDEsIlxcaW50cm9TIiwyXSxbMSwyLCJjIFxcOyBcXG1hcHN0byBcXEhvbV9DKC0sIGMpIiwyXSxbMiwwLCJQIFxcbWFwc3RvIFAoMSkiLDJdLFswLDQsIlxcaW50cm9OIDogXFxpZF9UIFxcdG8gXFxCb3hfVCIsMSx7InNob3J0ZW4iOnsidGFyZ2V0IjoyMH19XV0=
\[\begin{tikzcd}
	&& T \\
	\\
	{\Glob{C}} &&&& {\Glob{\Psh{C}}}
	\arrow["\introS"', from=1-3, to=3-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{c \; \mapsto \Hom_C(-, c)}"', from=3-1, to=3-5]
	\arrow["{P \mapsto P(1)}"', from=3-5, to=1-3]
	\arrow["{\introN : \id_T \to \Box_T}"{description}, shorten >=7pt, Rightarrow, from=1-3, to=0]
\end{tikzcd}\]

Keeping in mind \magicref{SliceBoxIsAspectBox},  this extends automatically to an indexed endofunctor on $T/-$, which we may call $\Box_{T/-}$, whose aspect at each $t \in T$ is given by the above diagram relative to the slice locally introspective theory $\slicePreIntrosp{T}{t}$. \sTODOinline{Remark on how these $\introN$ do indeed assemble into an indexed functor; i.e., the reindexing of $\introN$ at a globally defined object is the same as the slice introspective theory's $\introN$ at this object.} (We abuse notation slightly in the following diagram and write $\introN$ for the map from identity to $\Box_{T/-}$ whose each aspect is given by the appropriate instance of the previous diagram.)

% https://q.uiver.app/?q=WzAsMyxbMiwwLCJULy0iXSxbMCwyLCJDIl0sWzQsMiwiXFxQc2h7Q30iXSxbMCwxLCJcXGludHJvRiIsMl0sWzEsMiwiYyBcXDsgXFxtYXBzdG8gXFxIb21fQygtLCBjKSIsMl0sWzIsMCwiUCBcXG1hcHN0byBQKDEpIiwyXSxbMCw0LCJcXGludHJvTiA6IFxcaWRfe1QvLX0gXFx0byBcXEJveF97VC8tfSIsMSx7InNob3J0ZW4iOnsidGFyZ2V0IjoyMH19XV0=
\[\begin{tikzcd}
	&& {T/-} \\
	\\
	C &&&& {\Psh{C}}
	\arrow["\introF"', from=1-3, to=3-1]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{c \; \mapsto \Hom_C(-, c)}"', from=3-1, to=3-5]
	\arrow["{P \mapsto P(1)}"', from=3-5, to=1-3]
	\arrow["{\introN : \id_{T/-} \to \Box_{T/-}}"{description}, shorten >=7pt, Rightarrow, from=1-3, to=0]
\end{tikzcd}\]

We do not have such a strong natural transformation from identity to $\Box$ as acting on the other corners of the triangle ($C$ or $\Psh{C}$). However, by taking the natural transformation from $t$ to $\Box_T t$ and whiskering it on both sides along the trips from any other corner of the triangle into and out of $T$, we get a natural transformation from $\Box$ to $\Box \Box$ at each other corner of the triangle as well.

(The general principle here is that, given morphisms $f$ and $g$ between the same pair of objects in opposite directions, composing with $f$ on one side and $g$ on the other turns $(gf)^n$ into $(fg)^{n + 1}$, and thus whiskering in this way turns a natural transformation $: (gf)^n \to (gf)^m$ into a natural transformation $: (fg)^{n + 1} \to (fg)^{m + 1}$)

% https://q.uiver.app/?q=WzAsOCxbMSwwLCJcXFBzaHtDfSJdLFsyLDAsIlQvLSJdLFszLDEsIkMiXSxbNCwxLCJcXFBzaHtDfSJdLFs1LDAsIlQvLSJdLFszLDBdLFs2LDAsIkMiXSxbMCwwLCJDIl0sWzAsMV0sWzEsMl0sWzIsM10sWzMsNF0sWzEsNCwiXFxpZF9UIiwwLHsibGV2ZWwiOjIsInN0eWxlIjp7ImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNCw2XSxbMiw2LCJcXEJveF9DIiwyLHsiY3VydmUiOjV9XSxbNywwXSxbNyw2LCJcXEJveF9DIiwwLHsiY3VydmUiOi01fV0sWzcsMiwiXFxCb3hfQyIsMl0sWzEyLDEwLCJcXGludHJvTiIsMix7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH19XV0=
\[\begin{tikzcd}
	C & {\Psh{C}} & {T/-} & {} && {T/-} & C \\
	&&& C & {\Psh{C}}
	\arrow[from=1-2, to=1-3]
	\arrow[from=1-3, to=2-4]
	\arrow[""{name=0, anchor=center, inner sep=0}, from=2-4, to=2-5]
	\arrow[from=2-5, to=1-6]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{\id_T}", Rightarrow, no head, from=1-3, to=1-6]
	\arrow[from=1-6, to=1-7]
	\arrow["{\Box_C}"', curve={height=30pt}, from=2-4, to=1-7]
	\arrow[from=1-1, to=1-2]
	\arrow["{\Box_C}", curve={height=-30pt}, from=1-1, to=1-7]
	\arrow["{\Box_C}"', from=1-1, to=2-4]
	\arrow["\introN"', shorten <=4pt, shorten >=4pt, Rightarrow, from=1, to=0]
\end{tikzcd}\]

% https://q.uiver.app/?q=WzAsNyxbMCwwLCJcXFBzaHtDfSJdLFsxLDAsIlQvLSJdLFsyLDEsIkMiXSxbMywxLCJcXFBzaHtDfSJdLFs0LDAsIlQvLSJdLFsyLDBdLFs1LDAsIkMiXSxbMCwxXSxbMSwyXSxbMiwzXSxbMyw0XSxbMSw0LCJcXGlkX1QiLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDZdLFszLDYsIlxcQm94X3tcXFBzaHtDfX0iLDJdLFswLDMsIlxcQm94X3tcXFBzaHtDfX0iLDIseyJjdXJ2ZSI6NX1dLFswLDYsIlxcQm94X3tcXFBzaHtDfX0iLDAseyJjdXJ2ZSI6LTV9XSxbMTEsOSwiXFxpbnRyb04iLDIseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9fV1d
\[\begin{tikzcd}
	{\Psh{C}} & {T/-} & {} && {T/-} & C \\
	&& C & {\Psh{C}}
	\arrow[from=1-1, to=1-2]
	\arrow[from=1-2, to=2-3]
	\arrow[""{name=0, anchor=center, inner sep=0}, from=2-3, to=2-4]
	\arrow[from=2-4, to=1-5]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{\id_T}", Rightarrow, no head, from=1-2, to=1-5]
	\arrow[from=1-5, to=1-6]
	\arrow["{\Box_{\Psh{C}}}"', from=2-4, to=1-6]
	\arrow["{\Box_{\Psh{C}}}"', curve={height=30pt}, from=1-1, to=2-4]
	\arrow["{\Box_{\Psh{C}}}", curve={height=-30pt}, from=1-1, to=1-6]
	\arrow["\introN"', shorten <=4pt, shorten >=4pt, Rightarrow, from=1, to=0]
\end{tikzcd}\]

(In the above two diagrams, all parallel paths commute except where an explicit natural transformation is noted, and all unmarked edges are the corresponding of the three edges along our triangle. In the diagram illustrating axiom 4 for $\Box_{\Psh{C}}$, we note that we can think of the codomain of $\Box_{\Psh{C}}$ as either $\Psh{C}$ or more narrowly its full subcategory $C$.)

Thus, our $\Box$ follows all the rules of the modal logic K4, in each of these contexts. Shortly, we shall see that the general logic followed by $\Box_C$ in a locally introspective theory is conversely no stronger than K4, while in an introspective theory, it is the modal logic GL. Indeed, in the very next chapter we will see how in an introspective theory we get the last ingredient for the modal logic GL, \Loeb/'s theorem.

\sTODOinline{Citation to Boolos and other modal logic references at some point}

\sTODOinline{Add commentary on how box extends to a bifunctor, even without presuming cartesian closure. Add commentary on the relationship with double negation/Booleanness when Lob's theorem is introduced}

\subsection{As applied to our archetypal examples}

\subsubsection{ZF-Finite examples}\label{ZFFiniteModal}
Recall from \magicref{SigmaModelComplex} the introspective theory $\langle \ZfinSigma, \InnerZfin, \introS, \introN \rangle$, where $\ZfinSigma$ is the lexcategory of $\Sigma_1$-definable hereditarily finite sets and $\Sigma_1$-definable functions between them up to provable equivalence in ZF-Finite, $\InnerZfin$ is the lexcategory internal to $\ZfinSigma$ of arbitrary definable sets and arbitrary definable functions between them up to provable equivalence in ZF-Finite, $\introS$ assigns to each piece of $\ZfinSigma$ the corresponding (globally defined) piece of $\InnerZfin$, and $\introN$ is the $\Sigma_1$-definable function which sends any hereditarily finite set to the canonical term describing it, as well as witnessing the provable entailment from truth to provability for $\Sigma_1$ formulae.

Let us now consider the behavior of our $\Box$ operators here. For convenience of discussion, we will discuss just their global aspects.

First, observe that we have $\Box_{\ZfinSigma} : \ZfinSigma \to \ZfinSigma$, defined by $\Hom_{\InnerZfin}(1, \introS(-))$. This takes any object in $\ZfinSigma$ and sends it to the object of global sections of the corresponding object in $\ZfinSigma$'s internal view of itself via $\InnerZfin$. This is like sending \quote{the object of $X$es} to \quote{the object of definitions of $X$es within ZF-Finite}. In the particular case where the object in question is subterminal (thus representing a proposition), this is like sending the proposition $X$ to the proposition \quote{There is a proof in ZF-Finite of X}. Thus, this operator acts just as the traditional provability operator in this context.

Next, we turn our attention to the $\ZfinSigma$-internal endolexfunctor $\Box : \InnerZfin \to \InnerZfin$. Again, we shall just look at its global aspect. Presuming the soundness of ZF-Finite in our metatheory for further convenience, we have that $\Glob{\InnerZfin}$ is identified with $\Zfin$, the lexcategory of arbitrary definable sets and arbitrary definable functions between them up to provable equivalence in ZF-Finite. Thus, we are considering $\Box : \Zfin \to \Zfin$, defined by $\introS(\Hom_{\Zfin}(1, -))$. This again takes any object in $Z$ and sends it to the object of global sections of the corresponding object in $Z'$ (the internal view $Z$ has of itself). So again, in the same way, this is like sending \quote{the object of $X$es} to \quote{the object of $X$es which are definable within ZF-Finite}, and in the particular case where the object in question is subterminal and thus represents a proposition, this is like sending the proposition $X$ to the proposition \quote{There is a proof in ZF-Finite of X}. So again, this operator acts just as the traditional provability operator in this context.

\subsubsection{Kripke frame example}\label{KripkeFrameModal}
Recall the locally introspective theory $\langle \Psh{Q}, C \rangle$ from \magicref{KripkeLocallyIntrosp}, constructed from a transitive relation $<$ on a set $P$, with $Q$ being $P$ augmented with a new maximum element $\infty$ and construed as a preorder category using the $<$ relation. Our definition of $C$ was such that $\Glob{C} = \Psh{|P|} = \Set^{|P|}$. Our definition of $\introS : \Psh{Q} \to \Psh{|P|}$ was the forgetful functor which ignores the $<$ relation and $\infty$ element in $Q$. Our definition of $C$ also ensures that $\Hom_C(1, -) : \Psh{|P|} \to \Psh{Q}$ sent $f \in \Psh{|P|}$ to the presheaf $F$ on $Q$ such that $F(q)$ is the product of $f(p)$ over all $p < q$, with restriction maps given by projection.

Thus, $\Box_{\Psh{|P|}}$ acts like so: For $F \in \Psh{|P|}$ and $x \in P$, we have that $\Box_{\Psh{|P|}}(F)(x) = \introS(\Hom_C(1, F))$ is the product of $F(y)$ over all $y < x$.

Similarly, $\Box_{\Psh{Q}}$ acts like so: For $F \in \Psh{Q}$ and $x \in Q$, we have that $\Box_{\Psh{Q}}(F)(x)$ is the product of $F(y)$ over all $y < x$. (This $\Box_{\Psh{Q}}(F)$ furthermore is equipped with restriction maps given by projection.)

In particular, we may consider the special case of subterminal presheaves, amounting to propositions which may be true in some places while not in others. Subterminal presheaves in $\Psh{|P|}$ are simply subsets of $P$. Given such a subset $F \subseteq P$, the above tells us that $\Box_{\Psh{|P|}}(F)$ is identified with the (downwards closed) subset of $P$ containing precisely those $x$ such that all $y < x$ are contained in $F$.

Similarly, subterminal presheaves in $\Psh{Q}$ are simply downwards closed subsets of $Q$. Given such a downwards closed subset $F \subseteq Q$, the above tells us that $\Box_{\Psh{Q}}(F)$ is identified with the downwards closed subset of $Q$ containing precisely those $x$ such that all $y < x$ are contained in $F$.

Thus, our notion of $\Box$ acts on these subterminal presheaves in precisely the manner of the traditional interpretation of the $\Box$ operator in Kripke frames. 

In this way, we may note that any identity which holds of the $\Box_C$ operator in all locally introspective theories $\langle T, C \rangle$ must be an identity which holds of the $\Box$ operator on propositions in Kripke frames constructed from a transitive $<$ relation. This in turn is well known to entail holding of the $\Box$ operator in the modal logic K4 \sTODOinline{cite}. Thus, in this sense (which we refrain from stating formally for now, though it clearly could be turned into various formal statements), the logic of the $\Box$ operator in locally introspective theories is no stronger than the modal logic K4. We've also already seen in \magicref{ModalAxiom4Section} that it is at least as strong as the modal logic K4, and thus the logic of the $\Box$ operator in general locally introspective theories can be identified with K4.

In the above, we discussed the locally introspective theory constructed by \magicref{KripkeLocallyIntrosp} from a transitive relation $<$. Recall that, in the particular case where $<$ is furthermore well-founded, we may impose cardinality constraints in a suitable fashion to turn this into an introspective theory (this being \magicref{KripkeIntrosp}). The nature of the $\Box$ operators on the introspective theory \magicref{KripkeIntrosp} will be essentially the same as for \magicref{KripkeLocallyIntrosp}, as the former is a sub-introspection of the latter\sTODOinline{Discuss the interaction between $\Box$ and sub-introspections}. Thus, again, these $\Box$ operators will match the traditional interpretation of the box operator in Kripke frames. In this way, we find that the logic of the $\Box_C$ operators in introspective theories $\langle T, C \rangle$ is no stronger than that of the $\Box$ operator in well-founded transitive Kripke frames, which is well known to correspond to the modal logic GL (given by K4 + \Loeb/'s theorem). In the next chapter, we will prove the remarkable fact that \Loeb/'s theorem (and thus all of GL) is validated in all introspective theories, so that the logic of the $\Box_C$ operator in introspective theories can be said to be precisely that of the modal logic GL.

\subsubsection{Step-indexing example}\label{StepIndexingModal}
Recall the locally introspective theory $\langle \Psh{\omega}, C \rangle$ from \magicref{StepIndexingLocallyIntrosp}, in which $\introS : \Psh{\omega} \to \Glob{C}$ acts as an equivalence. It was also observed in that construction that $\Hom_C(1, \introS(-)) : \Psh{\omega} \to \Psh{\omega}$ acts as the operator $\Later$, such that $\Later(P)(0) = 1$ and $\Later(P)(n + 1) = P(n)$, for $P \in \Psh{\omega}$ and $n \in \omega$.

Thus, the operator $\Box_{\Psh{\omega}}$ here is precisely this operator $\Later$.

Note that this is quite distinct from the $\Box$ operator obtained for \magicref{KripkeLocallyIntrosp} as discussed above, even when considering the poset $\omega$. The distinction is that $\Later(P)(n)$, for positive $n$, is not the product of $P(m)$ at all $m < n$ but rather simply their limit, $P(n - 1)$.

However, this distinction is not visible when considering only subterminal presheaves in $\Psh{\omega}$. These amount to downwards closed subsets of $\omega$, and for such a downwards closed subset $F \subseteq \omega$, we will have that $\Box_{\Psh{\omega}}$ is the downwards closed subset of $\omega$ which includes $n$ just in case either $n = 0$ or $n - 1 \in F$, which in turn happens just in case all $m < n$ are contained in $F$. This again is the same behavior as the traditional $\Box$ operator on a Kripke frame corresponding to $\omega$. Note however that in this locally introspective theory, we do not have the ability to discuss propositions which are not downwards closed.

The $\Box$ operators on the introspective theory \magicref{StepIndexingIntrosp} are again essentially the same as those on the locally introspective theory \magicref{StepIndexingLocallyIntrosp}, as the former is a sub-introspection of the latter\sTODOinline{Again, discuss how sub-introspections relate to the $\Box$ operators}.

\sTODOinline{As applied to sub-introspections: If $i : S \to T$ is the inclusion of a full sublexcategory, and the pre-introspective theory $\langle S, \pullAlong{i} C \rangle$ is obtained from the locally introspective theory $\langle T, C \rangle$, then we get a map from $S \to \Psh{S}$ which acts morally like $\Box_S$ by $\Hom_{\pullAlong{i}C}(1, \introS(-))$, and this also matches the map we get by composing $i : S \to T$, with $\Box_T : T \to T$, with $\pullAlong{i} : T \to \Psh{S}$. In this sense, $S$ and $T$ agree on $\Box$. If $S$ is closed under $\Box_T$, we can put this more nicely.}

\subsection{Recap}
We have defined the $\Box$ operator in the context of locally introspective theories $\langle T, C \rangle$, acting on each of $T$, $C$, and $\Psh{C}$.

We have furthermore shown how each of these $\Box$ operators satisfies the rules of K4 modal logic.

We have also seen that the traditional interpretation of K4 modal logic in transitive Kripke frames is a special case of our $\Box$ operator for locally introspective theories (thus letting us conclude a tight correspondence between locally introspective theories and K4 modal logic).

The traditional provability operator on ZF-Finite (or just as well, on Peano Arithmetic) was also seen as a special case of our $\Box$ operator on an introspective theory. And the step-indexing operator considered in guarded recursion on the topos of trees was also seen as a special case of our $\Box$ operator on locally introspective or introspective theories.

In the next chapter, we will show that in any introspective theory, the $\Box$ furthermore satisfies the \Loeb/ property of the modal logic GL (aka, \quote{provability logic}). Thus, the correspondence between locally introspective theories and the modal logic K4 is complemented by a correspondence between introspective theories and the modal logic GL.
\fileend